The \tsc{r}-pronoun plus postposition \tit{waar mee} in Dutch

### 2021-11-16 paper.tex
ok, alleen nog drivation + conclusion

\documentclass[12pt]{article}

@@

\title{The \tsc{r}-pronoun and postposition \tit{waar-mee} in Dutch}
\author{Fenna Bergsma\\Goethe-Universität Frankfurt}
\date{\today}

@@

Dutch has the preposition \tit{met} `with' that expresses the instrumental. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{'t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{'t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{'t} `it' can be used as the object of a verb.

\ex.
\ag. Ik schilder met een kwast.\\
 I paint with a brush\\
 `I am painting with a brush.'\label{ex:metdp}
 \bg. Ik zie 't.\\
  I see it\\
  `I see it.'\label{ex:tverb}

\tit{Met} `with' and \tit{'t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{'r} `there' and the postposition \tit{mee} `with', as shown in \ref{ex:jarmee}.\footnote{The \tsc{r}-pronoun \tit{'r} in \ref{ex:jarmee} can be written as \tit{er}, \tit{der} and \tit{'r} and pronounced as respectively /ɛr/, /dər/ or /ər/. As far as I am aware, there is no clear meaning difference between these forms. See \citet{wesseling2018} for discussion.
In my examples I use \tit{'r}, but the other two forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

\ex.\label{ex:rmeeconst}
\ag. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}
\bg. Ik schilder 'r -mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:jarmee}

\tsc{R}-pronouns \citep{riemsdijk1978,koopman1994} are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r}. The adpositions they combine with obligatorily follow the \tsc{r}-pronoun, see \ref{ex:jarmee} and \ref{ex:neemeer}.
Notice also that the preposition \tit{met} `with' differs phonologically from the postposition \tit{mee} `with` (see \ref{ex:metdp} and \ref{ex:jarmee}).

\exg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:neemeer}

The main question I address in this paper is how to correctly rule out \tit{met 't} `with it' in \ref{ex:neemett}, and let the \tsc{r}-pronoun and postposition in \ref{ex:jarmee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout mechanisms. Just like \citet{riemsdijk1978} I analyze an \tsc{r}-pronoun and postposition as a type of allomorph of the preposition and pronoun. The crucial difference in the current approach is that spellout rules out the ungrammaticality of the preposition and pronoun and not the stipulation of a filter.

This paper is focuses on the instrumental \tsc{r}-pronoun and postposition \tit{'r-mee} `with it' and \tit{waar-mee} `with what' in Dutch. This instance is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative is intuitive, as many prepositions express locations, directions etc. However, an instrumental expresses an instrument, which does not have a meaning component associated with location. Ideally, an analysis treats the syncretism not as accidental but allows for the locative meaning to be absent.
The second reason why I focus on this particular \tsc{r}-pronoun and postposition has to do with the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{-mee} `with' when it is combined with an \tsc{r}-pronoun. This last observation has so far remained unexplained.

The main generalization is that the instrumental \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' takes precedence over the instrumental preposition and inanimate pronoun \tit{met wat} `with what'. This generalization is subject to an important condition: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition and pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, regular pronouns appear with prepositions, \tsc{r}-pronouns with postpositions. Third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I discuss the condition that \tsc{r}-pronouns and postpositions are subject too: they need to form a proper constituent. I show that \tsc{r}-pronouns and postpositions and prepositions and pronouns are in complementary distribution, depending on whether all features form a proper constituent or not.
In Section \ref{sec:takingapart} I decompose \tit{waar-mee} `with what' and \tit{met wat} `with what', and I connect them to parts of syntactic structure. Section \ref{sec:derivation} shows in a derivation how \tit{waar-mee} `with what' is derived and the role that constituency plays. Section \ref{sec:conclusion} concludes and discusses possible extensions of the analysis. All examples in this paper are from Dutch, unless indicated otherwise. Unmarked examples are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all relevant features form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{'rmee} `with it' and \tit{waar-mee} `with what` are the default, that normally they appear instead of \tit{met 't} `with it' and \tit{met wat} `with what`. Then I show an exception, in which \tit{met wat} `with what` has to be used and \tit{waar-mee} `with what` cannot. I argue that the crucial difference between the exception and other cases is that in the exception the relevant features do not form a proper constituent.

@@

In what follows I discuss the distribution of \tsc{r}-pronouns and regular pronouns in general \citep{riemsdijk1978,koopman1994}. I start with the personal pronouns and then return to the wh-pronouns.

Dutch has the accusative personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{'t} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

 \ex. \label{ex:objverb}
 \ag. Ik zie haar/hem.\\
  I see her/him\\
  `I see her/him.'\label{ex:aniobj}
 \bg. Ik zie 't.\\
  I see it\\
  `I see it.'\label{ex:inaniobj}

Example \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. Repeating from the introduction, the inanimate personal pronoun \tit{'t} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an \tsc{r}-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{ex:erprep} shows that the \tsc{r}-pronoun has moved obligatorily to the left of the adposition.

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met 't.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder 'r -mee.\\
 I paint there -with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee 'r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

\tit{Met} is not the only preposition with which this happens. For example, \tit{op} `on' and \tit{in} `in' also do not combine with the inanimate personal pronoun \tit{'t}, but the \tsc{r}-pronoun and postposition are used obligatorily.

\ex.
\ag. Ik zit 'r op.\\
 I sit there on\\
 `I am sitting on it.
\bg. *Ik zit op 't.\\
 I sit on it\\
 `I am sitting on it.

\ex.
 \ag. Hij zwemt 'r in.\\
  he swims it-in\\
  `He is swimming there it.'
 \bg. *Hij zwemt in 't.\\
  he swims in it\\
  `He is swimming in it.'

The situation of the inanimate wh-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as an object of a verb (see \ref{ex:wat}), but not as an object of a preposition \ref{ex:metwat}. In that case, the \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' appears, as shown in \ref{ex:waar-mee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar -mee schilder jij?\\
 where with paint you with\\
 `What are you painting with?'\label{ex:waar-mee}

\tit{Waar-mee} `with what' and not \tit{met wat} `with what' does not only appear in wh-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waar-mee} `with what' is used.

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waar -mee jij ook schildert.\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waar -mee jij ook schildert.\\
  I paint where with you also paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

In sum, \tit{'t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by respectively \tit{'r} `there' and \tit{waar} `where'.

\subsection{\tit{Met wat} `with what' shows up}

This section discusses an exception to what is shown in the previous section. I show an instance in which \tit{met wat} `with what' has to be used instead of \tit{waar-mee} `with what'. This instance comes from a mismatching free relative construction.

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the embedded clause) combine with two different cases (i.e. the case requirements do not match). I give an example in which the \tsc{r}-pronoun and postposition appear before I get to the exception. Consider \ref{ex:gekochtwaar-mee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun and postposition \tit{waar-mee} `with what' is used here. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.\footnote{In this example, \tit{waar} `where' takes \tit{-mee} `with' to the left edge of the embedded clause. It is also possible for \tit{-mee} `with' to be stranded, and \tit{waar} `where' to be moved to the left edge of the embedded clause on its own.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought waar you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
\z.

\phantom{x}
}

\ex.
\ag. Ik heb gekocht waar -mee jij schildert.\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaar-mee}
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

If the predicates are switched between the clauses, the \tsc{r}-pronoun and postposition do not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the embedded clause. The use of an \tsc{r}-pronoun and postposition is ungrammatical, as indicated by \ref{ex:schilderwaar-mee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{wat} `what' in used.

\ex.\label{ex:schildermet}
\ag. *Ik schilder waar -mee jij hebt gekocht.\\
 I paint where with you have bought\\
 `I paint with what you bought.'\label{ex:schilderwaar-mee}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:schildermetwat}

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waar-mee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waar-mee} `with what' is ungrammatical and \tit{met wat} `with what' is used.

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waar-mee} and \tit{met wat}}
	\begin{minipage}{0.45\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waar-mee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabularx}
\label{tbl:distribution}
\end{minipage}
\end{table}

In the remainder of this section I argue that the crucial point of \ref{ex:schildermetwat} is that the instrumental object does not form a proper constituent, i.e. it is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns and postpositions contain an instrumental object that does form a proper constituent.

Below I repeat the examples with instrumentals I discussed so far in this paper.

\ex.
\ag. Ik schilder 'r -mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar -mee schilder jij?\\
where with paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar -mee jij ook schildert].\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar -mee jij ook schildert].\\
 I paint where with you also paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

In each of these examples the instrumental object forms a proper constituent at a certain point in the derivation. In \ref{ex:const1}, the instrumental object forms a proper constituent in the surface order, as shown in \ref{ex:const1stage}. In \ref{ex:const2}, the instrumental object forms a proper constituent before wh- and V2- movement, shown in \ref{ex:const2stage}.
The structure in \ref{ex:const3stage} represents a stage in the derivation of the embedded clauses in \ref{ex:const3} and \ref{ex:const4}. Again, in the stage, which comes before relative movement of the pronoun to the left periphery of the relative clause, the instrumental object forms a proper constituent.

\ex.
\a. [[ik] [[schilder] ['r -mee]]]\label{ex:const1stage}
\b. [[jij] [[schilder] [waar-mee]]]\label{ex:const2stage}
\b. [[jij] [[ook] [[schilder] [waar-mee]]]]\label{ex:const3stage}

Let me now show how this applies to the examples with the mismatching free relatives. The two predicates I used in the free relatives are \tit{kopen} `to buy' and \tit{schilderen}  `to paint'. \tit{Kopen} `to buy' takes an accusative object, illustrated in \ref{ex:kopen}. \tit{Schilderen} `to paint' can take an instrumental as its object, shown in \ref{ex:schilderen}.\footnote{\tit{Schilderen} also optionally takes an (accusative) object, but I am focusing on the instrumental object here.}

@@

I repeat the mismatching free relative in which \tit{waar-mee} `with what' appears in \ref{ex:mismatchwaar-mee}. The predicate \tit{schildert} `paints' combines in the embedded clause with the instrumental object. The instrumental object forms a proper constituent within the embedded clause, and the it can be realized as the \tsc{r}-pronoun and postposition \tit{waar-mee} `with what'.\footnote{I assume that the accusative case requirement of \tit{gekocht} `bought' is satisfying by grafting a subconstituent of \tit{waar-mee} `with what' \citep{bergsma2019}.}

\exg. Ik heb gekocht [waar -mee jij schildert].\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaar-mee}

Next, I arrive at the mismatching free relative in which \tit{waar-mee} `with what' cannot be used, but \tit{met wat} `with what' appears. The embedded clause predicate \tit{gekocht} `bought' combines with an accusative object. The accusative object of a verb is always \tit{wat} `what', as I showed in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent, and \tit{waar-mee} `with what' does not surface.

@@

 The mismatching free relative in \ref{ex:mismatchmetwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watwasfur}, \tit{wat} `what' is the \tit{wat} `what' in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
 In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both constructions \tit{wat} `what' takes a complement and \tit{met wat} `with what' do not form a proper constituent. The brackets within the examples indicate the constituency.

@@

For the purpose of this paper I let \tit{a} correspond to \tsc{deixP}.

\ex. \begin{forest}
[\tsc{deixP}
    [\tsc{deix}, roof]
]
{\draw (.east) node[right]{⇔ \tit{a}}; }
\end{forest}\label{ex:entrya}

>>>>

\documentclass[11pt,a4paper]{article}

@@

\title{The \tsc{r}-pronoun plus postposition \tit{waar mee} in Dutch}
\author{Fenna Bergsma\\Goethe-Universität Frankfurt}
\date{\today}

@@

Dutch has the preposition \tit{met} `with' that expresses the instrumental. In \ref{ex:metdp}, \tit{met} `with' is combined with a full DP. The inanimate pronoun in Dutch is \tit{t} `it'.\footnote{The longer form that is mostly used in writing is \tit{het} `it'. I will use the spoken variant \tit{t} throughout this paper.} \ref{ex:tverb} illustrates that \tit{t} `it' can be used as the object of a verb.

\ex.
\ag. Ik schilder met een kwast.\\
 I paint with a brush\\
 `I am painting with a brush.'\label{ex:metdp}
 \bg. Ik zie t.\\
  I see it\\
  `I see it.'\label{ex:tverb}

\tit{Met} `with' and \tit{t} `it' do not appear together, as illustrated in \ref{ex:neemett}. Instead, Dutch uses the \tsc{r}-pronoun \tit{r} `there' and the postposition \tit{mee} `with', as shown in \ref{ex:jar mee}.\footnote{The \tsc{r}-pronoun \tit{r} in \ref{ex:jar mee} can be written as \tit{er}, \tit{der} and \tit{r} and pronounced as respectively /ɛr/, /dər/ or /ər/. There is no clear meaning difference between these forms that is relevant for the discussion in this paper
(see \citet{wesseling2018} for discussion).
In my examples I use \tit{r}, but the other forms fit just as well.}$^,$\footnote{In this paper I do not make any claims about the distinction between prefixes and prepositions, or suffixes and postpositions.}

\ex.\label{ex:r meeconst}
\ag. *Ik schilder met t.\\
 I paint with it\\
 `I am painting with it.'\label{ex:neemett}
\bg. Ik schilder r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:jar mee}

\tsc{R}-pronouns are nominal elements that are syncretic with locative pronouns, which in Dutch means they contain the morpheme \tit{r} \citep[cf.][]{riemsdijk1978,koopman1994}. The adpositions they combine with obligatorily follow the \tsc{r}-pronoun (compare \ref{ex:jar mee} and \ref{ex:neemeer}).
Notice also that the preposition \tit{met} `with' differs phonologically from the postposition \tit{mee} `with` (see \ref{ex:metdp} and \ref{ex:jar mee}), although they express the same meaning.

\exg. *Ik schilder mee r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:neemeer}

The main question I address in this paper is how to correctly rule out \tit{met t} `with it' in \ref{ex:neemett}, and let the \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in \ref{ex:jar mee} appear. I argue that \tsc{r}-pronouns are not something special, but a consequence of regular spellout mechanisms. Just as \citet{riemsdijk1978} I analyze the \tsc{r}-pronoun plus postposition as a type of allomorph of the preposition plus pronoun. The crucial difference in the current approach is that spellout - not the stipulation of a filter - rules out the ungrammaticality of the preposition plus inanimate pronoun.

This paper focuses on the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' in Dutch. This instance is interesting for two reasons. First, just as for all \tsc{r}-pronouns, the \tsc{r}-pronoun is syncretic with the locative. In many of the \tsc{r}-pronouns and postpositions, the meaning component of the locative within the \tsc{r}-pronoun plus postposition is intuitive, as many adpositions express locations, directions, etc. However, an instrumental expresses an instrument, which does not have a meaning component associated with location. Ideally, a formal analysis would treat the syncretism as unaccidental but still allow the locative meaning to be absent.
The second reason for focusing on this particular \tsc{r}-pronoun plus postposition is the form of the adposition. The preposition \tit{met} `with' does not only turn into a postposition, but it also changes into \tit{mee} `with' when it is combined with an \tsc{r}-pronoun. This last observation has so far remained unexplained in the literature on \tsc{r}-pronouns (cf. \citealt{koopman1994,riemsdijk1978,noonan2017dutch,abels2003diss}).

The main generalization is that the instrumental \tsc{r}-pronoun plus postposition \tit{r mee} `with it' takes precedence over the instrumental preposition plus inanimate pronoun \tit{met t} `with it'. I show that a single requirement has to be satisfied: the instrumental object needs to form a proper constituent i.e. a constituent to the exclusion of other features. When this condition is not met, the preposition plus inanimate pronoun appear. This can straightforwardly follow in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2009}. I work this idea out capturing the following three observations. First, \tsc{r}-pronouns are syncretic with locatives. Second, inanimate pronouns appear with prepositions, and \tsc{r}-pronouns appear with postpositions. And third, the instrumental preposition and postposition differ in form (\tit{met} vs. \tit{mee}).

This paper is structured as follows. In Section \ref{sec:distribution} I show that \tsc{r}-pronouns plus postpositions and prepositions plus inanimate pronouns are in complementary distribution, depending on whether all features form a proper constituent or not. Besides discussing \tit{r mee} `with it' and \tit{met t} `with it', I introduce the \tit{wh}-pronouns \tit{waarmee} `with what' and \tit{met wat} `with what'. For these forms it is intuitive to see that proper constituency is the critical condition.
In Section \ref{sec:takingapart} I decompose \tit{waar mee} `with what' and \tit{met wat} `with what', and I link the morphemes they consist of to specific syntactic structures. Section \ref{sec:derivation} shows in a derivation how \tit{waar mee} `with what' is derived when all features form a proper constituent. I also show how \tit{met wat} `with what' appears when the features do not form a proper constituent. Section \ref{sec:conclusion} concludes. All examples in this paper are from spoken Dutch, unless indicated otherwise. Examples without a reference are constructed and have been verified by native speakers.

@@

The goal of this section is to show that \tsc{r}-pronouns and postpositions appear when all relevant features form a proper constituent, i.e. that all features form a constituent to the exclusion of any other features. I start the section by showing that \tit{r mee} `with it' and \tit{waar mee} `with what` are the default, which is to say that they normally appear and \tit{met t} `with it' and \tit{met wat} `with what' do not. There is one exception, in which \tit{met wat} `with what` has to be used and \tit{waar mee} `with what` cannot, which is when case the relevant features do not form a proper constituent.

@@

In what follows I discuss the distribution of \tsc{r}-pronouns and regular pronouns in general \citep{riemsdijk1978,koopman1994}. I start with the personal pronouns and then turn to the \tsc{wh}-pronouns.

Dutch has the accusative personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{t} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

 \ex. \label{ex:objverb}
 \ag. Ik zie haar/hem.\\
  I see her/him\\
  `I see her/him.'\label{ex:aniobj}
 \bg. Ik zie t.\\
  I see it\\
  `I see it.'\label{ex:inaniobj}

Example \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. Repeating from the introduction, the inanimate pronoun \tit{t} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an \tsc{r}-pronoun appears, as is illustrated in \ref{ex:preper}. \ref{ex:erprep} shows that the \tsc{r}-pronoun must appear to the left of the adposition.

\ex. \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met t.\\
 I paint with it\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee r.\\
 I paint with there\\
 `I am painting with it.'\label{ex:erprep}

\tit{Met} is not the only preposition this occurs with. For example, \tit{op} `on' and \tit{in} `in' cannot combine with the inanimate pronoun \tit{t} either, and the \tsc{r}-pronoun plus postposition are obligatorily used instead.

\ex.
\ag. Ik zit r op.\\
 I sit there on\\
 `I am sitting on it.
\bg. *Ik zit op t.\\
 I sit on it\\
 `I am sitting on it.

\ex.
 \ag. Ik zwem r in.\\
  I swim there in\\
  `I am swimming in it.'
 \bg. *Ik zwemt in t.\\
  I swim in it\\
  `I am swimming in it.'

The situation of the inanimate \tsc{wh}-pronouns resembles the one of the inanimate personal pronouns. \tit{Wat} `what' can function as the object of a verb (see \ref{ex:wat}), but not as the object of a preposition \ref{ex:metwat}. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' appears, as shown in \ref{ex:waar mee}.\footnote{The sentence in \ref{ex:metwat} is unacceptable with neutral intonation. It becomes acceptable if \tit{wat} `what' is stressed, for example in a context in which the speaker is highly surprised about the choice of the object the hearer is painting with.}

\ex.
\ag. Wat zie jij?\\
 what see you\\
 `What do you see?'\label{ex:wat}
\bg. *Met wat schilder jij?\\
 with what paint you\\
 `What are you painting with?'\label{ex:metwat}
\bg. Waar mee schilder jij?\\
 where with paint you\\
 `What are you painting with?'\label{ex:waar mee}

\tit{Waar mee} `with what' and not \tit{met wat} `with what' does not only appear in \tsc{wh}-questions, but also in other contexts. \ref{ex:headed} gives an example of a headed relative, and \ref{ex:headless} shows a free relative in which both predicates combine with an instrumental object. The use of \tit{met wat} `with what' is ungrammatical in both contexts, and \tit{waar mee} `with what' is used instead.\footnote{
In this example, \tit{waar} `where' and \tit{mee} `with' are the two left-most elements of the relative clause. It is also possible for \tit{mee} `with' to appear more to the right in the clause and for \tit{waar} to appear as the left-most element on its own.

\exg. Ik schilder waar jij ook mee schildert.\\
 I paint where you also with paint\\
 `I am painting with what you are painting with too.'\label{ex:meealong}

I return to this variant in Section \ref{sec:derivation}.
}

\ex.\label{ex:headed}
\ag. Ik schilder met de kwast waar mee jij ook schildert.\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'
\bg. *Ik schilder met de kwast met wat jij ook schildert.\\
 I paint with the brush with what you also paint\\
 `I am painting with the brush that you are painting with too.'

 \ex.\label{ex:headless}
 \ag. Ik schilder waar mee jij ook schildert.\\
  I paint where with you also paint\\
  `I am painting with what you are painting with too.'
 \bg. *Ik schilder met wat jij ook schildert.\\
  I paint with what you also paint\\
  `I am painting with what you are painting with too.'

In sum, the inanimate pronouns \tit{t} `it' and \tit{wat} `what' do not combine with prepositions. They are substituted by, respectively, the \tsc{r}-pronouns \tit{r} `there' and \tit{waar} `where'.

\subsection{When \tit{met wat} `with what' shows up}

This section starts by showing that \tit{met wat} `with what' has to be used instead of \tit{waar mee} `with what' under certain circumstances in mismatching free relatives. Later in this section I argue that under these circumstances the relevant features that need to be spelled out do not form a proper constituent.

A mismatching free relative is a free relative construction in which the two predicates (the one in the main clause and the one in the relative clause) combine with two different cases. In other words, the case requirements of the two clauses do not match. First consider \ref{ex:gekochtwaarmee}. The predicate in the relative clause, \tit{schildert} `paint', combines with an instrumental object. The predicate in the main clause, \tit{gekocht} `bought' combines with an accusative object. The \tsc{r}-pronoun plus postposition \tit{waar mee} `with what' is used here, shown in \ref{ex:gekochtwaarmee}. The use of \tit{met wat} `with what' is ungrammatical in this context, illustrated in \ref{ex:gekochtmetwat}.

\ex.
\ag. Ik heb gekocht waar mee jij schildert.\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

If the predicates are switched between the clauses, the \tsc{r}-pronoun plus postposition cannot appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental object in the main clause and \tit{gekocht} `bought' combines with an accusative object in the relative clause.
The use of an \tsc{r}-pronoun plus postposition is ungrammatical, illustrated in \ref{ex:schilderwaarmee}. Instead, a combination of the instrumental preposition \tit{met} `with' and the \tsc{wh}-pronoun \tit{wat} `what' is used, shown in \ref{ex:schildermetwat}.

\ex.\label{ex:schildermet}
\ag. *Ik schilder waar mee jij hebt gekocht.\\
 I paint where with you have bought\\
 `I paint with what you bought.'\label{ex:schilderwaarmee}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:schildermetwat}

Table \ref{tbl:distribution} summarizes the pattern. When the main clause predicate combines with an accusative and the relative clause predicate with an instrumental, \tit{waar mee} `with what' is grammatical and \tit{met wat} `with what' is ungrammatical. When the main clause predicate combines with an instrumental and the relative clause predicate with an accusative, \tit{waar mee} `with what' is ungrammatical and \tit{met wat} `with what' appears instead.

\begin{table}[ht]
  \center
	\caption {Distribution between \tit{waar mee} and \tit{met wat}}
		\begin{tabular}{lll}
		\toprule
                              & \tit{waar mee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, r:\tsc{ins}  & ✔             & *             \\
    m:\tsc{ins}, r:\tsc{acc}  & *             & ✔             \\
    \bottomrule
\end{tabular}
\label{tbl:distribution}
\end{table}

In the remainder of this section I argue that the crucial point of \ref{ex:schildermetwat} is that the instrumental object does not form a proper constituent, i.e. the instrumental object is not a constituent to the exclusion of any other elements. The other side of the coin is that constructions with \tsc{r}-pronouns and postpositions contain an instrumental object that does form a proper constituent.

Below I repeat the examples with instrumental objects discussed in this paper so far. I placed square brackets around the relative clauses in \ref{ex:const3} and \ref{ex:const4}. In each of these examples the instrumental object forms a proper constituent at a certain point in the derivation.

\ex.
\ag. Ik schilder r mee.\\
 I paint there with\\
 `I am painting with it.'\label{ex:const1}
\bg. Waar mee schilder jij?\\
where with paint you with\\
 `What are you painting with?'\label{ex:const2}
\bg. Ik schilder met de kwast [waar mee jij ook schildert].\\
 I paint with the brush where with you also paint\\
 `I am painting with the brush that you are painting with too.'\label{ex:const3}
\bg. Ik schilder [waar mee jij ook schildert].\\
 I paint where with you also paint\\
 `I am painting with what you are painting with too.'\label{ex:const4}

Consider now the examples with the mismatching free relatives. The two predicates I used in the free relatives are \tit{kopen} `to buy' and \tit{schilderen} `to paint'. \tit{Kopen} `to buy' takes accusative objects, illustrated in \ref{ex:kopen}. \tit{Schilderen} `to paint' takes instrumental objects, shown in \ref{ex:schilderen}.\footnote{\tit{Schilderen} also optionally takes an (accusative) object, but I am focusing on the instrumental object here.}

@@

In \ref{ex:mismatchwaarmee}, I repeat the mismatching free relative in which \tit{waar mee} `with what' appears. The predicate \tit{schildert} `paints' combines in the relative clause with the instrumental object. I placed square brackets around the relative clauses in \ref{ex:mismatchwaarmee}. The instrumental object forms a proper constituent within the relative clause, and it can be realized as the \tsc{r}-pronoun plus postposition \tit{waar mee} `with what'.

\exg. Ik heb gekocht [waar mee jij schildert].\\
 I have bought where with you paint\\
 `I bought what you are painting with.'\label{ex:mismatchwaarmee}

Consider now the other kind of mismatch. The relative clause predicate \tit{gekocht} `bought' combines with an accusative object. I placed square brackets around the relative clauses in \ref{ex:mismatchmetwat}. The accusative \tsc{wh}-object of a verb is always \tit{wat} `what', as I have shown earlier in \ref{ex:wat}. The instrumental only comes into the picture in the main clause, when \tit{schilder} `paint' combines with an instrumental object. At no point in the derivation does the instrumental object form a proper constituent. \tit{Waar mee} `with what' cannot surface, and \tit{met wat} appears instead.

@@

 The mismatching free relative in \ref{ex:mismatchmetwat} is not the only construction in which the string \tit{met wat} `with what' appears. I give examples of two more occurrences in \ref{ex:moremetwat}. In \ref{ex:watwasfur}, \tit{wat} `what' is the \tit{wat} in the so-called \tit{wat voor} `what for'-construction \citep[cf.][]{corver1991}.
 In \ref{ex:watindef}, \tit{wat} appears as a quantifier, and it means `some'. In both constructions \tit{wat} `what' takes a complement and \tit{met wat} `with what' is not a proper constituent. The brackets within the examples indicate the constituency.

@@

For the purpose of this paper I let \tit{a(a)} correspond to \tsc{deixP}.

\ex. \begin{forest}
[\tsc{deixP}
    [\tsc{deix}, roof]
]
{\draw (.east) node[right]{⇔ \tit{a(a)}}; }
\end{forest}\label{ex:entrya}
